Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-24 main.tex
Edits to the introduction.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll rate~\footnote{Roll angular rate is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle at lower speeds. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the high speed self-stability
already does. We have developed a balance assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single-track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so solely with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode~\footnote{The
motorcycle (and bicycle) eigenmodes are defined in \citep{Sharp1971}.}. They
also proposed how the gains must change with respect to vehicle speed for
favorable control at all speeds.  This roll motion feedback enables the
simplest controller that can stabilize a single-track vehicle above a minimum
speed when one is not concerned with wobble instabilities. Ruijs and Pacejka's
work was not particularly concerned with low speed stability and their vehicle
was fully automatic, i.e no human rider was involved.

Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single-actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to effectively change
the dynamics, up to some limits, of the human-controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist system but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is suitable for showing the effect of a roll
motion feedback driven steer motor on the dynamics. This model is equivalently
valid for on-road or treadmill motion~\citep{Kooijman2009a}, which have the
same dynamics. The linear version of this model~\citep{Meijaard2007} can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) matrix is a function of the equilibrium
forward speed \(v\). It and the input \(\mathbf{B}\) matrix are otherwise
populated with expressions that are functions of the geometric and inertial
parameters of the nonholonomic multibody system made up of four rigid bodies:
two wheels, front frame, and rear frame.

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:
%
\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability for a low speed range. We use this simple
feedback principle as the basis for our balance assist controller.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of their necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll angular rate~\footnote{Roll angular
rate is defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle down to very low speeds. If automatic control of steering
can stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citep{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \cite{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, by roll
angular rate feedback stabilizes the weave mode, and by steer angular rate
feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
gains must change with respect to vehicle speed for favorable control across
all speeds. This roll motion feedback enables the simplest controller that can
stabilize a single-track vehicle above a minimum speed when one is not
concerned with wobble instabilities. But Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, i.e no human rider was involved.

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act on the steer in parallel. The effect of this automatic control gives the
ability to effectively change the human-controlled plant dynamics, up to some
limits. In our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced
motion during distractions due to the balance assist system but
\citet{Hanakam2023} recently showed rider dissatisfaction with the
stabilization of a similar vehicle so the possible benefits are not
definitively established.

The design of a balance assist system relies on the linear Carvallo-Whipple
bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits non-minimum phase behavior and self-stability. The bicycle
model is well suited for showing the effect of a roll motion feedback driven
steer motor on the dynamics. This model is equivalently valid for on-road or
treadmill riding~\citep{Kooijman2009a}, which have the same fundamental
dynamics. The linear version of this model can be described by the fourth order
state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and the motor torque is a proportional feedback controller
\(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then the human-controlled
plant takes the form:
%
\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means the control gain \(k_{\dot{\phi}}\) can be selected such that the
eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. By gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborate on some of the possibilities in scheduling the
gains for such a controller for a bicycle and shows that a linear scheduling
with respect to speed can give satisfactory stability for a low speed range. We
use this simple feedback principle as the basis for our balance assist
controller.
